---
title: "Infant Mortality, Liberalizations and Interventionism: A Causal Analysis Accounting for Data Quality"
subtitle: "Data-quality proxy: `rp` (relative precision)"
date: today
author: "Geloso & Wilhelm"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
link-external-newwindow: true
---

```{r setup}
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)

# Load pre-computed results from 2) Analysis.R
qd <- readRDS("analysis_results.rds")
invisible(list2env(qd, envir = environment()))

# Which data-quality proxy this document is built around
PROXY <- "rp"             # "con_int" or "rp"

# Significance star helper
sig_stars <- function(p) {
  case_when(p < 0.01 ~ "***", p < 0.05 ~ "**", p < 0.10 ~ "*", TRUE ~ "")
}
```

# Introduction

This analysis tests whether large, sustained changes in economic freedom
(EFW) cause changes in infant mortality (IMR), and — more pointedly —
**how much of that effect is hidden or revealed when we control for
data quality**.

The treatment of interest is a 5-year EFW score change of $\geq +1$ point
("liberalization") or $\leq -1$ point ("deliberalization" / restriction).
The outcome is the 5-year change in the country's infant mortality rate.
Methodology mirrors Callais & Young (2023) and Grier (2025), with three
matching estimators per test (PSM NN3, PSM Kernel, Mahalanobis NN3 with
Abadie-Imbens bias adjustment).

We run **8 tests** combining: treatment direction, presence/absence of a
data-quality control, and presence/absence of a control-group filter that
removes opposite-direction events from the pool of potential matches.

**Two data-quality proxies** are examined in parallel — the analysis
script (`2) Analysis.R`) computes results for both, and this document
reports **`rp` (relative precision)**. A companion document presents the same
analysis using `con_int` (absolute CI width).

::: {.callout-note}
## Tradeoffs between `con_int` and `rp` as a data-quality control

Both measures derive from the UNICEF/UN-IGME Bayesian B-spline model and
quantify the uncertainty around each country-year IMR estimate.

- **`con_int`** = $\text{IM\_upper} - \text{IM\_lower}$ — the absolute
  width of the 90% credible interval. *Pros:* directly interpretable in
  IMR units (deaths per 1,000); strongly correlated with sample size and
  vital-registration completeness. *Cons:* mechanically larger when IMR
  itself is larger, so high-mortality countries look "lower quality" by
  default even if their measurement infrastructure is identical.
- **`rp`** = $(\text{con\_int}/2) / \text{IMR}$ — relative precision: the
  half-width as a fraction of the point estimate. *Pros:* scale-free;
  isolates measurement quality from the level of mortality. *Cons:*
  becomes unstable when IMR is very low (denominator approaches zero),
  potentially exaggerating "noise" in advanced economies that are
  actually well-measured.

Vincent's preference is **`con_int`**: easier to interpret, and the
mechanical correlation with mortality level is partially absorbed by the
lagged IMR covariate (`lag_IM`) that is already in the propensity-score
model. The `rp`-based analysis serves as a robustness check.
:::


# Summary Statistics

```{r sumstats-table}
sumstats %>%
  kbl(caption = "Summary statistics — full estimation sample",
      col.names = c("Variable", "Mean", "Std. Dev.", "Min", "Max"),
      digits = 3, booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, position = "left",
                bootstrap_options = c("striped", "hover", "condensed")) %>%
  pack_rows("Outcome variable",     1, 1) %>%
  pack_rows("Data-quality proxies", 2, 3) %>%
  pack_rows("EFW change variable",  4, 4) %>%
  pack_rows("PSM covariates",       5, 15)
```


# Treatment Definitions and Episodes

We identify two treatment events in a 5-year quinquennial panel from
1970 to 2015. Both are constructed in `1) Import_Merge.R`.

| Treatment | Definition | Interpretation |
|:----------|:-----------|:---------------|
| **EFW Score Jump** | EFW index rose $\geq +1.0$ point over 5 years | **Liberalization** |
| **EFW Score Drop** | EFW index fell $\leq -1.0$ point over 5 years | **Deliberalization** (restriction) |

Observations adjacent to any qualifying event within the same country are
set to NA (lag 1, lag 2, lead 1) to avoid contamination from overlapping
reform episodes (Callais & Young 2023). Venezuela 2000 is excluded as a
structural break.

::: {.callout-note}
## Rank-based treatments removed

The previous version of this analysis (archived in `Code/Archive/`) also
examined EFW *rank* changes as alternative treatments. Per the 5/8/2026
direction, the rank analysis has been scrapped from the current paper.
:::

## EFW Score Jumps ($\Delta\text{EFW} \geq +1.0$)

```{r tbl-jump}
tbl_EFWjump %>%
  kbl(caption = paste0("EFW Score Jump episodes (N = ", nrow(tbl_EFWjump), "). ",
                       "Sorted by magnitude of EFW increase."),
      col.names = c("Country", "Period", "EFW Change"),
      booktabs = TRUE, align = c("l","c","r")) %>%
  kable_styling(full_width = FALSE, position = "left",
                bootstrap_options = c("striped","hover","condensed"))
```

## EFW Score Drops ($\Delta\text{EFW} \leq -1.0$)

```{r tbl-drop}
tbl_EFWdrop %>%
  kbl(caption = paste0("EFW Score Drop episodes (N = ", nrow(tbl_EFWdrop), "). ",
                       "Sorted by magnitude of EFW decrease."),
      col.names = c("Country", "Period", "EFW Change"),
      booktabs = TRUE, align = c("l","c","r")) %>%
  kable_styling(full_width = FALSE, position = "left",
                bootstrap_options = c("striped","hover","condensed"))
```


# PSM Methodology

## Three Matching Estimators

For each of the 8 tests we estimate the Average Treatment Effect on the
Treated (ATT) using three matching specifications. Differences across
the estimators are a robustness check; agreement strengthens the causal
interpretation.

### 1. PSM — 3 Nearest Neighbors (NN3)
Logit propensity score → 3 nearest neighbors on the score, caliper of
0.25 SDs, no replacement. Bootstrap SEs (200 reps, seed 12345).

### 2. PSM — Epanechnikov Kernel
Same logit PS as NN3; each treated unit compared to a kernel-weighted
average of all controls within the caliper. Bootstrap SEs (200 reps).

### 3. Mahalanobis NN3 with Bias Adjustment
Skips the PS model. Each treated unit matched to its 3 nearest controls
in the full covariate space using Mahalanobis distance, with
Abadie-Imbens (2011) bias correction. Analytic SEs.

## Covariates

All covariates are **lagged one 5-year period** (pre-treatment levels).
Covariates marked **(*)** appear only in tests where data quality is
included as a control:

| Covariate | Description | Source |
|:----------|:------------|:-------|
| `lagEFW`            | EFW summary index (lagged)         | Fraser Institute |
| `laghc`             | Human capital index                | Penn World Tables |
| `laglngdppc`        | Log GDP per capita                 | Penn World Tables |
| `laglngdppc2`       | Log GDP per capita squared         | Penn World Tables |
| `laggdpc_5growth`   | 5-year GDP per capita growth       | Penn World Tables |
| `lagfertilrate`     | Fertility rate                     | World Bank WDI |
| `lagoldagedep`      | Old-age dependency ratio           | World Bank WDI |
| `lagpolity2`        | Polity2 democracy score            | Polity5 |
| `lagurbanpop`       | Urban population share             | World Bank WDI |
| `lag_IM`            | Lagged infant mortality rate       | UNICEF |
| ``lag_rp`` **(*)** | Lagged relative precision | UNICEF |


# Results

For each test we report ATT, SE, p-value and 95% CI from the three
matching estimators. Bootstrap SE (200 reps) for PSM NN3 and PSM Kernel;
Abadie-Imbens analytic SE for Mahalanobis. Stars: \* p<0.10, \*\* p<0.05,
\*\*\* p<0.01.

A **negative ATT** indicates the treated group experienced a *larger
drop* in IMR than matched controls — beneficial for liberalizations,
adverse for deliberalizations.

```{r results-loop, results="asis"}
res_for_proxy <- results_by_proxy[[PROXY]]

# Render a single test's result table
render_test <- function(tspec, res) {
  cat(sprintf("\n## %s\n\n", tspec$label))
  if (is.null(res)) {
    cat("Insufficient treated units after filtering — skipped.\n\n")
    return(invisible(NULL))
  }
  res_fmt <- res %>%
    mutate(
      ATT_str = sprintf("%.4f%s", ATT, sig_stars(p_value)),
      SE_str  = sprintf("(%.4f)", SE),
      CI_str  = sprintf("[%.4f, %.4f]", CI_lo, CI_hi)
    ) %>%
    dplyr::select(Estimator, `N Treated` = N_treated,
                  ATT = ATT_str, SE = SE_str, `95% CI` = CI_str)
  print(kbl(res_fmt,
            caption = tspec$label, booktabs = TRUE, escape = FALSE,
            align = c("l","r","r","r","r")) %>%
          kable_styling(full_width = FALSE, position = "left",
                        bootstrap_options = c("striped","hover","condensed")) %>%
          footnote(general = "Bootstrap SE (200 reps) for PSM NN3 and PSM Kernel; Abadie-Imbens SE for Mahalanobis. *** p<0.01, ** p<0.05, * p<0.10.",
                   general_title = "Notes:", footnote_as_chunk = TRUE))
  cat("\n\n")
}

# --- Panel A: Liberalization (1pt jumps) ---
cat("## Panel A — Liberalization (1pt EFW Jump) {.unnumbered}\n\n")
for (n in 1:4) {
  tspec <- test_grid[[n]]
  render_test(tspec, res_for_proxy[[as.character(n)]])
}

# --- Panel B: Deliberalization (1pt drops) ---
cat("## Panel B — Deliberalization (1pt EFW Drop) {.unnumbered}\n\n")
for (n in 5:8) {
  tspec <- test_grid[[n]]
  render_test(tspec, res_for_proxy[[as.character(n)]])
}
```


# Balance & Distribution Diagnostics

For each test, we show the propensity-score density (before vs after
PSM NN3 matching) and a Love plot of standardized mean differences for
every covariate in that test's PS model. Reference lines at $\pm 0.10$
(Austin 2011) mark the conventional well-balanced threshold.

```{r balance-loop, results="asis", fig.width=9, fig.height=4.8}
plots_for_proxy <- balance_plots[[PROXY]]

for (n in 1:8) {
  tspec <- test_grid[[n]]
  plots <- plots_for_proxy[[as.character(n)]]
  cat(sprintf("\n## %s {.unnumbered}\n\n", tspec$label))
  if (is.null(plots)) {
    cat("Balance plots unavailable (insufficient treated units).\n\n")
    next
  }
  print(plots$ps_density); cat("\n")
  print(plots$love);       cat("\n\n")
}
```


# Caveats

::: {.callout-warning}
## Caveats

- **Years covered.** The current sample spans **1970–2015**, dictated by
  the latest year for which the UNICEF infant mortality estimates and
  the pre-built Callais & Young covariate panel both have coverage.
  Extending through 2023/24 will require sourcing more recent vintages
  of the underlying covariate series.
- **EFW & IM data.** The Fraser Institute EFW Index 2025 edition (1970–
  2023) and the UNICEF Infant Mortality Rate dataset used here are the
  latest publicly available vintages.
- **Covariates.** The propensity-score covariate set is taken **directly
  from Callais & Young (2023)**. To exploit the full 1970–2023 EFW
  window these covariates need to be **updated and extended through
  2023/24** (and potentially expanded to include health-specific
  controls relevant to infant mortality).
- **Data-quality proxy.** This document uses **`rp` (relative precision)**.
  Results using `con_int` (absolute CI width) appear in the companion
  document; differences across the two are part of the robustness
  story.
:::


# Questions for Vincent

*[To be filled in by EHW.]*

1.
2.
3.